Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project

### 2014-03-21 paper.tex
incorporated John Bent's comments

\begin{abstract}
With phase 1 of the Fast Forward Storage and IO Stack project complete, it is
an excellent opportunity to evaluate many of the decisions made to feed into
the phase 2 effort. The initial effort to define a next generation file system
has made admirable contributions in architecture and design. Formalizing the
general idea of data staging as burst buffers for the file system will help
manage the performance variability and offer the data processing opportunities
outside the main compute and file system. Adding a transactional mechanism to
manage faults and data visibility helps enable effective analytics without
having to work around the IO stack semantics.

While these and their other contributions are valuable, similar efforts made
elsewhere may offer attractive alternatives or differing semantics that could
yield a more feature rich environment with little to no additional overhead.
For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
alternative approach for incorporating transactional semantics into the data
path. The PreDatA project examined how to get the best throughput for data
operators can offer some insights for refining the Burst Buffer concept.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, a joint effort between the US Department of Energy's
Office of Advanced Simulation and Computing and Advanced Scientific Computing
Research commissioned an effort to develop a design and prototype for an IO
stack suitable for the extreme scale environment. This is a joint effort led by
Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL. Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
getting underway. The core focus of the first phase was basic functionality and
design. Overriding many of the decisions during this and any subsequent phases
is the reality of budgets. Placing GBs of NVRAM on every node, while a
potentially advantageous approach, is not financially feasible. With this in
mind, the second phase will refine this design incorporating fault recovery and
other features missing from the first phase.

The overall design seeks to offer a byte-granular, multi-version concurrency
control. Through the use of a copy-on-write style mechanism, multiple versions
of a file can be stored in potentially greatly reduced space. It assumes the
client interface will be through an IO library affording a more complicated
interface that offers richer functionality without requiring user code changes.
Managing most data access in a machine-local layer rather than requiring
writing to centralized storage will better support the performance and energy
requirements of extreme scale integrated application workflows. At a more
detailed view, the various layers of the IO stack each contribute different
functionality and performance implications.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library. The intent is to only have
access to the storage stack through such an API to manage the complexity of
working with the lower layers. Below the user API is an IO forwarding layer
that redirects IO calls from the compute nodes to the IO dispatching layer.
This IO forwarding layer is analogous to the function of the IO nodes in a
BlueGene machine. The next two layers have considerable functionality. The IO
dispatcher (IOD) serves as the primary storage interface for the IO stack and
is the only way to reach the persistent storage array in lower layers. The
Data Access Object Storage (DAOS) layer serves as the persistent storage
interface. At the bottom is the Versioning Object Storage Device (VOSD). It
serves as the interface for storing objects of all types efficiently.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
FastForward IOD design recommends incorporating SSDs, but specifically lists
these devices as optional. Unfortunately, both incorporating burst buffers and
the use of SSDs in the IOD layer may be problematic.  First, the IOD and DAOS
designs currently are written assuming burst buffers. This means that the
bulk of IO operations will only hit the IOD layer and proposed functionality,
such as the function operators, do not discuss the impact on the DAOS layer or
the operators themselves should a burst buffer not be available.  Consider the
important functionality of data rearrangement and doing things like changing
the fast array dimension on shared, spinning media.  The design of DAOS assumes
that it will only be involved when persisting a completed transaction and only
for a fraction of the total transactions created.

@@

In the case of the written IOD design, it describes a fixed-sized staging area
shared across all users simultaneously.  This is unlikely to be useful because
of the limited compute and communication capacity to spare to perform these
operations at a bottleneck in the IO path.  The use of a separate data staging
area intentionally separate from the IO path allows using operators on limited
resources leaving the IO path clear for strictly data movement. A nuance of
this design is discussed in the Design Philosophy below.

By concentrating this functionality into the storage stack, three problems
arise.  First, the amount of network bandwidth, IO bandwidth, and compute power
consumed for example operations from a single application is likely to
completely monopolize the IOD processes. Second, if space and time partitioning
is used instead, the functionality risks being too small to be useful. Third,
the hardware performance advantage for SSDs is questionable. Current NAND-based
flash devices top out at around 400 MB/sec. The key spec that is missing from
this number is that 400 MB/sec is a measure of the fixed number of available
IOPS multiplied by the block size. This represents the ideal streaming
performance possible. The problem is that it costs an IOP to read 1 byte or 1
block (4KB or 8KB, depending on the device). It costs 1 IOP to write a full
block--usually. In some cases, it will cost 2 IOPs.  This accounts for the
required pre-erase write prior to writing to a reused block. In the worst case,
it can be 3 IOPS per write. This would be for a partial block write (read the
old block, erase, write the modified block). One-third of 400 MB/sec, about 133
MB/sec, is well below the streaming performance of HDDs.  Granted, there is
still rotational and seek latency to deal with for HDDs, but the advantage for
SSDs has evaporated and potentially turned into a penalty at a considerable
cost premium.  There are faster SSD solutions on the market that incorporate
DRAM for caching and using the PCIe bus, for example, but their price precludes
them from use in an extreme scale platform.

@@

\subsection{Design Philosophy}
The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSDs for buffers to the IO nodes. The
team does acknowledge the limitations of this design and intend to ultimately
focus on spreading the IOD layer from the IO nodes into the compute area as
well.  This is intended both to help address the limitations of the IO
bandwidth and compute capability of these few nodes for data processing, but
also to take advantage of new layers in the storage hierarchy. By incorporating
NVRAM into compute nodes, new options for buffering data prior to being moved
to centralized storage become available and addresses some of the concerns
about SSD performance. This lessens the impact of some operators while offering
additional options for places to store data.

Burst buffers being optional is a high level goal, but not one considered at a
detailed level with the design. For example, if there is no burst buffer, all
of the advanced functionality proposed for the IOD layer would have to work
against the DAOS layer instead. For example, function shipping assumes it will
operate on fast, local data within the IOD layer rather than against the
globally shared DAOS layer. With the additional desire to support using compute
node resources for these operations, serious work will be required to make a
fully functional end-to-end IOD layer implementation for a production system.

Another concern that is acknowledged, but no thought has been applied to is the
requirement that a single IOD process be the master for any operation. Should
the number of concurrent applications exceed the available nodes, sharing an
IOD process will be required. Since partitioning of the IOD processes for
exclusive use by particular applications is the assumed operating mode, should
insufficient IOD resources be available, either a job could be delayed or IOD
resources could be reallocated from a different process could be redeployed for
use by the new job. These sorts of considerations still need to be made for a
full production system.

@@

The transaction mechanism manifests in two forms. At the IOD layer, they are
called transactions and are used to judge whether or not an output is complete
or not and control access through treating the transaction ID as a version
identifier. At the DAOS layer, they are called epochs and represent persisted
(durable) transactions from the IOD layer. Each of these offers different
functionality, but are connected as is explained below. How these differ from
the \DDT approach is also explored. While IOD's and \DDTns's transactions are
seemingly very different, they use a similar high-level design, but very
different implementation, to solve the same problem.

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container serves to host a transaction and contains a collection of
objects. Conceptually, containers corresponds to a file in a traditional file
system. The objects in each container represent different data within a file.
The three initially defined object types are key-value store, array, and blob.
The easiest way to understand these types is to evaluate these from the
perspective of an HDF-5 file, the initial user interface layer. The key-value
store represents a collection of attributes or groups. The array represents
part of a potentially multi-dimensional array. It is referred to as
a shard because it is likely a small piece of a globally defined array. The
blob represents a byte stream of arbitrary contents.  The fundamental
difference between an array and a blob is that the array has metadata
identifying its portion within the global, logical space while the blob is
simply a 1-dimensional array of bytes that is not shared across IOD nodes.
Should an operation be deployed into IOD to manipulate an array within a
container, it would operate on the array rather than on blobs. Given this
context, the transactions come in two forms.

@@

\subsection{DAOS Epochs}
The Epoch mechanism differs from transactions. Instead of focusing on when a
particular output is complete, an epoch represents incremental persisted copies
of a container. To simplify the mapping between an IOD transaction and the DAOS
epochs, when an IOD transaction is persisted to DAOS, the IOD transaction ID is
the used as the epoch ID. The key difference is that at the DAOS layer, some
transaction (epoch) IDs will not be represented.

@@

With a client-side only approach, there is a serious risk of the metadata
service and the object store getting out of sync.  While having a metadata-less
object storage service is desirable, the different semantics from traditional
file systems requires some considerations. In this case, should these services
get out of sync, three particular risks are introduced.  First, a client could
create an entry in the metadata service that does not correspond to any objects
in the object store. Second, a client could create objects that have no
associated metadata entries. Third, updates to the metadata or object store
service should be an atomic operation, but due to a lack of coordination, a
window where the system is inconsistent appears.

@@

At a broader level, there are some concerns that were partially clarified
through conversations with the team.  Consider a shared file system across an
HPC data center. The current design maintains the metadata in its own
container. Since copying data from the IOD layer to the DAOS layer requires an
explicit persist call, how and when synchronizing the metadata across the
layers and potentially across machines occurs is undefined. Delaying
synchronization until an explicit persist is called will reduce the update
frequency, but delays the data visibility. Ideally, the metadata object would
need to be automatically persisted every time a container transaction is
persisted to the DAOS layer.

@@

A point of confusion rather than a potential design challenge is the change in
definitions between the IOD layer and the DAOS layer.  For the IOD layer, a
container is a collection of objects. For the DAOS layer, a container is a
collection of shards. For the IOD an object may be a shard of a global array.
For DAOS, a shard can host a set of DAOS objects. Having the same names with
locally correct, globally conflicting definitions serves to confuse how the
system should work.

>>>>

\begin{abstract}
With phase 1 of the Fast Forward Storage and IO Stack project complete, it is
an excellent opportunity to evaluate many of the decisions made to feed into
the phase 2 effort. The initial effort to define a next generation storage
system has made admirable contributions in architecture and design. Formalizing
the general idea of data staging into burst buffers for the storage system will
help manage the performance variability and offer additional data processing
opportunities outside the main compute and storage system. Adding a
transactional mechanism to manage faults and data visibility helps enable
effective analytics without having to work around the IO stack semantics.

While these and other contributions are valuable, similar efforts made
elsewhere may offer attractive alternatives or differing semantics that could
yield a more feature rich environment with little to no additional overhead.
For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
alternative approach for incorporating transactional semantics into the data
path. Another project, PreDatA, examined how to get the best throughput for
data operators and may offer additional insights into further refinements of
the Burst Buffer concept.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, a joint effort between the US Department of Energy's
Office of Advanced Simulation and Computing and Advanced Scientific Computing
Research commissioned an effort to develop a design and prototype for an IO
stack suitable for the extreme scale environment. This is a joint effort led by
Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL. Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. Overriding many of the decisions during
this and any subsequent phases is the reality of budgets. Placing GBs of NVRAM
on every node, while a potentially advantageous approach, is not financially
feasible. With this in mind, the second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

The overall design seeks to offer byte-granular, multi-version concurrency
control. Through the use of a copy-on-write style mechanism, multiple versions
of an object can be stored in potentially greatly reduced space. It assumes the
client interface will be through an IO library affording a more complicated
interface that offers richer functionality requiring only minimal end-user code
changes.  Managing most data access in a platform-local layer rather than
requiring
writing to centralized storage will better support the performance and energy
requirements of extreme scale integrated application workflows. At a more
detailed view, the various layers of the IO stack each contribute different
functionality and performance implications.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is to only have
access to the storage stack through such an API to manage the complexity of
working with the lower layers. Below the user API is an IO forwarding layer
that redirects IO calls from the compute nodes to the IO dispatching layer.
This IO forwarding layer is analogous to the function of the IO nodes in a
BlueGene machine. The next two layers have considerable functionality. The IO
dispatcher (IOD) serves as the primary storage interface for the IO stack and
is the only way to reach the persistent storage array in lower layers. The
Data Access Object Storage (DAOS) layer serves as the persistent storage
interface. At the bottom is the Versioning Object Storage Device (VOSD). It
serves as the interface for storing objects of all types efficiently.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
The current FastForward IOD design recommends incorporating SSDs, but
specifically lists these devices as optional. Unfortunately, not incorporating
burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
the IOD and DAOS designs currently are written assuming burst buffers. This
means that the bulk of IO operations will only hit the IOD layer and proposed
functionality, such as the function shipping, do not discuss the impact on the
DAOS layer or the operators themselves should a burst buffer not be available.
Consider the important functionality of data rearrangement and doing things
like changing the fast array dimension on shared, spinning media.  The design
of DAOS assumes that it will only be involved when persisting a completed
transaction and only for a fraction of the total transactions created.
Transaction frequency at the IOD layer can be higher since IOD does not
``flatten'' transactions. Flattening is the process of making each transaction
fully-independent by duplicating any data unchanged from previous transactions.
DAOS epochs are persisted, flattened transactions from the IOD layer.

@@

In the case of the written IOD design, it describes a fixed-sized staging area
that is partitioned on a per-application basis.  This is unlikely to be useful
because of the limited compute and communication capacity to spare to perform
these operations at a bottleneck in the IO path.  The use of a separate data
staging area intentionally separate from the IO path allows using operators on
limited resources leaving the IO path clear for strictly data movement. A
nuance of this design is discussed in the Design Philosophy below.

By concentrating the Burst Buffers and function shipping into the storage
stack, three problems arise.  First, the amount of network bandwidth, IO
bandwidth, and compute power consumed for example operations from a single
application is likely to completely monopolize the IOD processes. Second, if
space and time partitioning is used instead, the functionality risks being too
small to be useful. Third, the hardware performance advantage for SSDs is
questionable. Current NAND-based flash devices top out at around 400 MB/sec.
The key spec that is missing from this number is that 400 MB/sec is a measure
of the fixed number of available IOPS multiplied by the block size. This
represents the ideal streaming performance possible. The problem is that it
costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
IOPs.  This accounts for the required pre-erase write prior to writing to a
reused block. In the worst case, it can be 3 IOPS per write. This would be for
a partial block write (read the old block, erase, write the modified block).
One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
performance of HDDs.  Granted, there is still rotational and seek latency to
deal with for HDDs, but the advantage for SSDs has evaporated and potentially
turned into a penalty at a considerable cost premium.  There are faster SSD
solutions on the market that incorporate DRAM for caching and using the PCIe
bus, for example, but their price precludes them from use in an extreme scale
platform.

@@

\subsection{Design Philosophy}
The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSD buffers to the IO nodes. The team
does acknowledge the limitations of this design and intend to ultimately focus
on spreading the IOD layer from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth and
compute capability of these few nodes for data processing, but also to take
advantage of new layers in the storage hierarchy. By incorporating NVRAM into
compute nodes, new options for buffering data prior to being moved to
centralized storage become available and addresses some of the concerns about
SSD performance. This lessens the impact of some operators while offering
additional options for places to store data.

Burst buffers being optional is a high level goal, but not one considered at a
detailed level within the design. For example, if there is no burst buffer, all
of the advanced functionality proposed for the IOD layer would have to work
against the DAOS layer instead. For example, function shipping assumes it will
operate on fast, local data within the IOD layer rather than against the
globally shared DAOS layer. With the additional desire to support using compute
node resources for these operations, serious work will be required to make a
fully functional end-to-end IOD layer implementation for a production system.

Another concern that is acknowledged, but no thought has been applied to, is
the requirement that a single IOD process be the master for any operation.
Should the number of concurrent applications exceed the available nodes,
sharing an IOD process will be required. Since partitioning of the IOD
processes for exclusive use by particular applications is the assumed operating
mode, should insufficient IOD resources be available, either a job could be
delayed or IOD resources could be reallocated from a different process could be
redeployed for use by the new job. Handling resilience concerns for the IOD
processes must also be address. These sorts of considerations still need to be
made for a full production system.

@@

The transaction mechanism manifests in two forms. At the IOD layer, they are
called transactions and are used to judge whether or not a set of distributed,
asynchronous modifications across a set of related objects is complete or not
and control access through treating the transaction ID of committed transaction
as a version identifier.  At the DAOS layer, they are called epochs and
represent persisted (durable) transactions from the IOD layer. Each of these
offers different functionality, but are connected as is explained below. How
these differ from the \DDT approach is also explored. While IOD's and \DDTns's
transactions are seemingly very different, they use a similar high-level
design, but very different implementation, to solve the same problem.

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container serves to host a transaction and contains a collection of
objects. Conceptually, containers corresponds to a file in a traditional file
system. The objects in each container represent different data within a file.
The three initially defined object types are key-value stores,
multi-dimensional arrays, and blobs.  The easiest way to understand these types
is to evaluate these from the perspective of an HDF-5 file, the initial user
interface layer. The key-value store represents a collection of attributes or
groups. The array represents part of a potentially multi-dimensional array. It
is referred to as a shard because it is likely a small piece of a globally
defined array. The blob represents a byte stream of arbitrary contents.  The
fundamental difference between an array and a blob is that the array has
metadata identifying its portion within the global, logical space while the
blob is simply a 1-dimensional array of bytes that is not shared across IOD
nodes.  Should an operation be deployed into IOD to manipulate an array within
a container, it would operate on the array rather than on blobs. Given this
context, the transactions come in two forms.

@@

\subsection{DAOS Epochs}
The Epoch mechanism differs from transactions. Instead of focusing on when a
particular output is complete, an epoch represents incremental persisted copies
of a container. To simplify the mapping between an IOD transaction and the DAOS
epochs, when an IOD transaction is persisted to DAOS, the IOD transaction ID is
the used as the epoch ID. The key difference is that at the DAOS layer, some
transaction (epoch) IDs will not be represented since not all IOD transactions
are necessarily persisted.

@@

With a client-side only approach, there is a serious risk of the metadata
service and the object store getting out of sync.  While having a metadata-less
object storage service is desirable, the different semantics from traditional
file systems requires some considerations. In this case, should these services
get out of sync, three particular risks are introduced.  First, a client could
create a dangling entry in the metadata service that does not correspond to any
objects in the object store. Second, a client could create orphaned objects
that have no associated metadata entries. Third, updates to the metadata or
object store service should be an atomic operation, but due to a lack of
coordination, a window where the system is inconsistent appears.

@@

At a broader level, there are some concerns that were partially clarified
through conversations with the team.  Consider a shared file system across an
HPC data center. The current design maintains the metadata in its own
container. Since copying data from the IOD layer to the DAOS layer requires an
explicit persist call, how and when synchronizing the metadata across the
layers and potentially across machines occurs is undefined. Delaying
synchronization until an explicit persist is called will reduce the update
frequency, but delays the data visibility on other platforms. Ideally, the
metadata object would need to be automatically persisted every time a container
transaction is persisted to the DAOS layer.

@@

A point of confusion rather than a potential design challenge is the change in
definitions between the IOD layer and the DAOS layer.  For the IOD layer, a
container is a collection of objects. For the DAOS layer, a container is a
collection of objects across a set of shards. For the IOD an object may be a
shard of a global array.  For DAOS, a shard can host a set of DAOS objects.
Having the same names with locally correct, globally conflicting definitions
serves to confuse how the system should work.
